---
title: "iCodon"
output: rmarkdown::html_vignette
author: Santiago G. Medina-Muñoz
vignette: >
  %\VignetteIndexEntry{iCodonR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This documentation describes three use cases of this package.

1. Compute the codon frequencies for a set of coding DNA sequences.
2. Predict mRNA stability of a set of coding sequences using the model developed by Medina et al. 2020.
3. Optimize mRNA stability with synonymous mutations.

```{r setup, message=FALSE, warning=FALSE}
library(iCodon)
library(dplyr)
```


## 1. Compute the codon frequencies for a set of coding DNA sequences

Suppose you have a table containing a set of coding DNA sequences like the one shown next.

````{r, echo=FALSE}
set.seed(12)

codin_seqs <- training %>% 
  sample_n(100) %>% 
  select(gene_id, coding) %>% 
  filter(!duplicated(coding))

````

```{r}
codin_seqs
```

If you wish to compute the codon-frequencies you can use the function `add_codon_counts`.

```{r}
 add_codon_counts(codin_seqs)
```

see the help `?add_codon_counts` for more information about this function.

## 2. Predict mRNA stability for a coding sequence

You can use the function `predict_stability` to predict the mRNA stability based on the codon sequence (Medina et al 2020).

Positive values indicate stability, while negative values indicate instability.

You need to specify a species. Possible species are:

- *Homo sapiens* (human)
- *Mus musculus* (mouse)
- *Danio rerio* (fish)
- *Xenopus laevis* (xenopus)


```{r}
predictor_human <- predict_stability(specie = "human")
# Now you can pass any coding sequence to this function
predictor_human("ATGTGGAGCGGCGGAGCTGAGCAACAACACCCTAAAACCGACAAATCTCACCGATGCAATGGCGTCGACAGCTCAAGAAGAAAGAACAGATCGCAGCGGTGGCGATATGAAGTCAAGAAAACTGGATGA")
```

If you have a table (see first example) you can add a column with the prediction
for each sequence.

```{r}
codin_seqs$prediction_optimality <- predictor_human(codin_seqs$coding)

select(codin_seqs, gene_id, prediction_optimality)
```

## 3. Optimize mRNA stability with synonymous mutations.


To optimize mRNA stability you can use the function `optimizer`.


```{r}
a_seq <- "ATGTGGAGCGGCGGAGCTGAGCAACAACACCCTAAAACCGACAAATCTCACCGATGCAATGGCGTCGACAGCTCAAGAAGAAAGAACAGATCGCAGCGGTGGCGATATGAAGTCAAGAAAACTGGATGA"

optimizer(
  sequence_to_optimize = a_seq,
  specie = "human", n_iterations = 10,
  make_more_optimal = TRUE,
  mutation_Rate = .1,
  max_abs_val = 2,
  n_Daughters = 4
)
```


See the help `?optimizer` for information about the parameters.
